---
title: "Automated Detection of Seasonal Epidemic Burden Levels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automated Detection of Seasonal Epidemic Burden Levels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(aedseo)
library(mem)
```

## Summary

To provide a concise overview of how the `seasonal_burden_levels()` algorithm operates, we utilize the same example data presented
in the `vignette("aedseo")`. The plot below illustrates the two `method` arguments available in the
`combined_seasonal_output()` function:

- **`intensity_levels`**: This method assesses burden levels by comparing it to observations from previous seasons.
- **`peak_levels`**: This method assesses burden levels by referencing only the highest-rate observations within each season.

The disease-specific threshold is the `very low` threshold for both methods.

```{r, echo = FALSE}
withr::local_seed(222)
# Construct an 'tsd' object with time series data
tsd_data_clean <- generate_seasonal_data(
  years = 3,
  start_date = as.Date("2021-10-18"),
  amplitude = 2000,
  phase = 0,
  time_interval = "week"
)

# Run models
intensity_levels <- seasonal_burden_levels(
  tsd = tsd_data_clean,
  disease_threshold = 140,
  method = "intensity_levels",
  conf_levels = 0.95
)
peak_levels <- seasonal_burden_levels(
  tsd = tsd_data_clean,
  disease_threshold = 140,
  method = "peak_levels",
  conf_levels = c(0.4, 0.9, 0.975),
  n_peak = 8
)

# Create data frame
burden_levels_df <- data.frame(
  Level = names(
    c(intensity_levels$values,
      peak_levels$values
    )
  ),
  Threshold = c(
    intensity_levels$values,
    peak_levels$values
  ),
  Method = c(
    rep("Intensity Levels", 4),
    rep("Peak Levels", 4)
  )
)
burden_levels_df$Level <- factor(
  burden_levels_df$Level,
  levels = c("high", "medium", "low", "very low")
)

# Create the plot
burden_levels_df |>
  ggplot2::ggplot(ggplot2::aes(x = 0, y = Threshold, linetype = Level, color = Level)) +
  ggplot2::geom_hline(
    ggplot2::aes(yintercept = Threshold, linetype = Level, color = Level),
    linewidth = 1
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Observations",
    linetype = "Aedseo levels",
    color = "Aedseo levels"
  ) +
  ggplot2::scale_linetype_manual(
    values = c(
      "very low" = "dotted",
      "low" = "dashed",
      "medium" = "longdash",
      "high" = "solid"
    )
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "very low" = "#44ce1b",
      "low" = "#bbdb44",
      "medium" = "#f2a134",
      "high" = "#e51f1f"
    )
  ) +
  ggplot2::facet_wrap(~ Method, ncol = 2) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    legend.position = "right",
    legend.key.width = grid::unit(2, "cm")
  ) +
  ggplot2::scale_y_log10(
    labels = scales::label_comma(big.mark = ".", decimal.mark = ",")
  )
```

## Methodology

The methodology used to define the burden levels of seasonal epidemics is based on observations from previous seasons.
Historical data from all available seasons is used to establish the levels for the current season.
This is done by:

- Using `n` peak weekly observations from each season.
- Selecting only peak observations if they surpass the disease-specific threshold.
- Weightening the observations such recent observations have a greater impact than older observations.
- A proper distribution (log-normal, weibull and exponential are implemented) is used to fit the weighted
 `n` peak observations. Then the parameters of the selected distribution are optimised to select the best fit.
- Burden levels can be defined by two methods:
  - `intensity_levels` which models the risk compared to what has been observed in previous seasons.
  - `peak_levels` which models the risk compared to what has been observed in the `n` peak observations each season.
    This is the method used in [MEM](https://github.com/lozalojo/mem), with log normal distribution and without weights.

The model is implemented in the `seasonal_burden_levels()` function of the `aedseo` package.
In the following sections we will describe the arguments for the function and how the model is build.

#### Peak observations
`n_peak` observations are used to describe the highest observations that are observed each season.
The default of `n_peak` is `6` as we are only interested in the highest observations.

#### Weightening
`decay_factor` is implemented due to more recent seasons often are more indicative of current and future trends.
 As time progresses, the relevance of older seasons may decrease due to changes in factors like population immunity,
 virus mutations, or intervention strategies. Weighting older seasons less reflects this reduced relevance.
The default of `decay_factor` is `0.8`, allowing the model to be responsive to recent changes without being overly
sensitive to short-term fluctuations.
The optimal decay factor can vary depending on the variability and trends within the data. For datasets where seasonal
patterns are highly stable, a higher decay factor might be appropriate. Conversely, data that has changed a lot across
seasons, a lower factor could improve predictions.

#### Distribution and optimisation
`family` is the argument used to select which distribution the `n_peak` observations should be fitted to, users can
choose between `lnorm`, `weibull` and `exp` distributions. The log-normal distribution theoretically
aligns well with the nature of epidemic data, which often exhibits multiplicative growth patterns.
In our optimization process, we evaluated the distributions to determine their performance in fitting danish non-sentinel
cases and hospitalisation data for RSV, SARS-CoV-2 and Influenza (A and B). All three distributions had comparable
objective function values during optimisation, hence we did not see any statistical significant difference in their performance.

The model uses the `fit_quantiles()` function which employes the `stats::optim` for optimisation of the distribution parameters.
The `optim_method` argument can be passed to `seasonal_burden_levels()`, default is `Nelder-Mead` but other methods can be selected,
see `?fit_quantiles`.

*Note:* [MEM](https://github.com/lozalojo/mem) uses the log-normal distribution, which allows for more straightforward benchmarking,
due to this the default is `lnorm`.

#### Burden levels
`method` is the argument used to select one of the two methods `intensity_levels`(default) and `peak_levels`.
Both methods return quantile(s) from the fitted distribution which are used to define the burden levels.
Burden levels are "very low", "low", "medium" and "high".

- `intensity_levels` takes one quantile as argument, representing the highest intensity that has been observed in previous seasons.
  The default is set at a 95% confidence level, which is used to determine the "high" burden level. The disease-specific threshold
  determines the "very low" burden level. The "low" and "medium" burden levels are calculated based on the relative
  increase between "very low" and "high" burden levels.

- `peak_levels` takes three quantiles as argument, representing the "low", "medium" and "high" burden levels.
  The default thresholds are set at 40%, 90%, and 97.5% to align with the parameters used in the [MEM](https://github.com/lozalojo/mem).
  The disease-specific threshold defines the "very low" burden level.

## Applying the `seasonal_burden_levels()` algorithm

To apply the `seasonal_burden_levels()` algorithm, data needs to be transformed into a `tsd` object.
```{r, include = FALSE}
withr::local_seed(222)
# Construct 'tsd' objects with time series data
tsd_data_noise <- generate_seasonal_data(
  years = 4,
  start_date = as.Date("2021-10-18"),
  amplitude = 2000,
  phase = 0,
  noise_sd = 200,
  time_interval = "week"
)

tsd_data_noise_and_pos_trend <- generate_seasonal_data(
  years = 4,
  start_date = as.Date("2021-10-18"),
  amplitude = 2000,
  phase = 0,
  noise_sd = 200,
  trend_rate = 1.002,
  time_interval = "week"
)

tsd_data_noise_and_neg_trend <- generate_seasonal_data(
  years = 4,
  start_date = as.Date("2021-10-18"),
  amplitude = 2000,
  phase = 0,
  noise_sd = 200,
  trend_rate = 0.99,
  time_interval = "week"
)

# Remove days after week 20 in last season to get only 4 seasons data
tsd_data_noise <- tsd_data_noise |>
  dplyr::filter(time <= "2025-05-12")
tsd_data_noise_and_pos_trend <- tsd_data_noise_and_pos_trend |>
  dplyr::filter(time <= "2025-05-12")
tsd_data_noise_and_neg_trend <- tsd_data_noise_and_neg_trend |>
  dplyr::filter(time <= "2025-05-12")

```

### Use the `intensity_levels` method
```{r}
intensity_levels_n <- seasonal_burden_levels(
  tsd = tsd_data_noise,
  disease_threshold = 140,
  method = "intensity_levels",
  conf_levels = 0.95
)
print(intensity_levels_n)
```

```{r, include = FALSE}
intensity_levels_n_pos_t <- seasonal_burden_levels(
  tsd = tsd_data_noise_and_pos_trend,
  disease_threshold = 140,
  method = "intensity_levels",
  conf_levels = 0.95
)
intensity_levels_n_neg_t <- seasonal_burden_levels(
  tsd = tsd_data_noise_and_neg_trend,
  disease_threshold = 140,
  method = "intensity_levels",
  conf_levels = 0.95
)
```

### Use the `peak_levels` method
[MEM](https://github.com/lozalojo/mem) uses the n highest values of each epidemic period to fit the parameters of the distribution,
where `n = 30/seasons`. The data has four seasons, to align with MEM, we use `n_peak = 8`
```{r}
peak_levels_n <- seasonal_burden_levels(
  tsd = tsd_data_noise,
  disease_threshold = 140,
  method = "peak_levels",
  conf_levels = c(0.4, 0.9, 0.975),
  n_peak = 8
)
print(peak_levels_n)
```

```{r, include = FALSE}
peak_levels_n_pos_t <- seasonal_burden_levels(
  tsd = tsd_data_noise_and_pos_trend,
  disease_threshold = 140,
  method = "peak_levels",
  conf_levels = c(0.4, 0.9, 0.975),
  n_peak = 8
)

peak_levels_n_neg_t <- seasonal_burden_levels(
  tsd = tsd_data_noise_and_neg_trend,
  disease_threshold = 140,
  method = "peak_levels",
  conf_levels = c(0.4, 0.9, 0.975),
  n_peak = 8
)
```

### Use the [Moving Epidemic Method (MEM)](https://github.com/lozalojo/mem)
MEM is run with default arguments.
```{r, include = FALSE}
data_list <- list(
  tsd_data_noise = tsd_data_noise,
  tsd_data_noise_and_pos_trend = tsd_data_noise_and_pos_trend,
  tsd_data_noise_and_neg_trend = tsd_data_noise_and_neg_trend
)
```

```{r}
# Run MEM algorithm
mem_thresholds <- purrr::map(data_list, ~ {
  mem_data <- .x |>
    dplyr::mutate(season = aedseo::epi_calendar(time),
                  week = lubridate::isoweek(time)) |>
    dplyr::select(-time) |>
    tidyr::pivot_wider(names_from = season, values_from = observation) |>
    dplyr::select(-week)
  # Run MEM
  mem_result <- mem::memmodel(mem_data, i.seasons = 4)
  # Extract thresholds
  mem_thresholds <- tibble::tibble(
    `epidemic threshold \n (start)` = mem_result$epidemic.thresholds[1],
    `epidemic threshold \n (end)` = mem_result$epidemic.thresholds[2],
    `medium` = mem_result$intensity.thresholds[1],
    `high` = mem_result$intensity.thresholds[2],
    `very high` = mem_result$intensity.thresholds[3]
  )
})
```
| Name                      | Value |
|---------------------------|-------|
| epidemic threshold (start)| 3004  |
| epidemic threshold (end)  | 2975  |
| medium                    | 3902  |
| high                      | 4206  |
| very high                 | 4347  |

## Compare `intensity_levels`, `peak_levels` and `mem` algorithms
The same data is created with following combinations:

- noise
- noise and positive trend
- noise and negative trend

These combinations are selected as it is realistic for real world data to have noise, and differentiation between
trend can occur declining or inclining between seasons.
Burden levels correspond to season *2024/2025* calculated based on the three previous seasons.

### Aedseo levels
```{r, echo = FALSE, fig.width=10, fig.height=8, dpi=300}
#### Create data frames
burden_levels_df <- tibble::tibble(
  Level = names(
    c(intensity_levels_n$values,
      intensity_levels_n_pos_t$values,
      intensity_levels_n_neg_t$values,
      peak_levels_n$values,
      peak_levels_n_pos_t$values,
      peak_levels_n_neg_t$values
    )
  ),
  Threshold = c(
    intensity_levels_n$values,
    intensity_levels_n_pos_t$values,
    intensity_levels_n_neg_t$values,
    peak_levels_n$values,
    peak_levels_n_pos_t$values,
    peak_levels_n_neg_t$values
  ),
  Method = c(
    rep("Intensity Levels \n (noise)", 4),
    rep("Intensity Levels \n (noise and \n positive trend)", 4),
    rep("Intensity Levels \n (noise and \n negative trend)", 4),
    rep("Peak Levels \n (noise)", 4),
    rep("Peak Levels \n (noise and \n positive trend)", 4),
    rep("Peak Levels \n (noise and \n negative trend)", 4)
  )
)

##### Adjustments for plot
burden_levels_df <- burden_levels_df |>
  dplyr::mutate(algorithm = stringr::str_extract(Method, ".*Levels"))

burden_levels_df$Level <- factor(
  burden_levels_df$Level,
  levels = c(
    "high",
    "medium",
    "low",
    "very low"
  )
)

# Calculate y_tics
y_tics_log10 <- pretty(c(log10(burden_levels_df$Threshold)))
y_tics_levels <- 10^(y_tics_log10)

# For each tic, find the closest magnitude to round correctly
round_to_nearest <- function(x) {
  magnitude <- 10^floor(log10(x))
  plyr::round_any(x, accuracy = magnitude)
}
y_tics <- sapply(y_tics_levels, round_to_nearest)

# Round max value
rounded_max_threshold <- round(max(burden_levels_df$Threshold) / 1000) * 1000
y_tics <- c(y_tics[-5], rounded_max_threshold, rounded_max_threshold + 2000)

#### Plot
burden_levels_df |>
  ggplot2::ggplot(ggplot2::aes(x = 0, y = Threshold, linetype = Level, color = Level)) +
  ggplot2::geom_hline(
    ggplot2::aes(yintercept = Threshold, linetype = Level, color = Level),
    linewidth = 1
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Observations",
    linetype = "Aedseo levels",
    color = "Aedseo levels"
  ) +
  ggplot2::scale_linetype_manual(
    values = c(
      "very low" = "dotted",
      "low" = "dashed",
      "medium" = "longdash",
      "high" = "solid"
    )
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "very low" = "#44ce1b",
      "low" = "#bbdb44",
      "medium" = "#f2a134",
      "high" = "#e51f1f"
    )
  ) +
  ggplot2::facet_wrap(~ Method, nrow = 2) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    legend.position = "right",
    legend.key.width = grid::unit(2, "cm")
  ) +
  ggplot2::scale_y_log10(
    breaks = y_tics,
    limits = range(y_tics),
    labels = scales::label_comma(big.mark = ".", decimal.mark = ",")
  )
```

### MEM levels
```{r, echo = FALSE, fig.width=10, fig.height=4}
mem_levels_df <- tibble::tibble(
  Level = names(
    c(mem_thresholds$tsd_data_noise,
      mem_thresholds$tsd_data_noise_and_pos_trend,
      mem_thresholds$tsd_data_noise_and_neg_trend
    )
  ),
  Threshold = c(
    as.numeric(mem_thresholds$tsd_data_noise),
    as.numeric(mem_thresholds$tsd_data_noise_and_pos_trend),
    as.numeric(mem_thresholds$tsd_data_noise_and_neg_trend)
  ),
  Method = c(
    rep("MEM Levels \n (noise)", 5),
    rep("MEM Levels \n (noise and \n positive trend)", 5),
    rep("MEM Levels \n (noise and \n negative trend)", 5)
  )
)

# Adjustments for plot
mem_levels_df$Level <- factor(
  mem_levels_df$Level,
  levels = c(
    "very high",
    "high",
    "medium",
    "epidemic threshold \n (start)",
    "epidemic threshold \n (end)"
  )
)

mem_levels_df |>
  ggplot2::ggplot(ggplot2::aes(x = 0, y = Threshold, linetype = Level, color = Level)) +
  ggplot2::geom_hline(
    ggplot2::aes(yintercept = Threshold, linetype = Level, color = Level),
    linewidth = 1
  ) +
  ggplot2::labs(
    x = NULL,
    y = "Observations",
    linetype = "MEM level",
    color = "MEM level"
  ) +
  ggplot2::scale_linetype_manual(
    values = c(
      "medium" = "longdash",
      "high" = "solid",
      "very high" = "dotdash",
      "epidemic threshold \n (start)" = "solid",
      "epidemic threshold \n (end)" = "solid"
    )
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "medium" = "#f2a134",
      "high" = "#e51f1f",
      "very high" = "#891212",
      "epidemic threshold \n (start)" = "#bbdb44",
      "epidemic threshold \n (end)" = "#44ce1b"
    )
  ) +
  ggplot2::facet_wrap(~ Method, ncol = 6) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank(),
    legend.position = "right",
    legend.key.width = grid::unit(2, "cm")
  ) +
  ggplot2::scale_y_log10(
    breaks = y_tics,
    limits = range(y_tics),
    labels = scales::label_comma(big.mark = ".", decimal.mark = ",")
  )
```

Upon examining all methods and data combinations, it becomes clear that the `intensity_levels` approach establishes 
levels covering the entire set of observations from previous seasons. In contrast, the `peak_levels` and `MEM` methods 
define levels solely based on the highest-rate observations within each season.

The highest observations for the *2024/2025* season for each data set are:

| Data                      | Value |
|---------------------------|-------|
| Noise                     | 4456  |
| Noise and positive trend  | 6048  |
| Noise and negative trend  | 3735  |

In relation to these highest observations and upon further examination, we observe the following:

1. Plots with Noise and Noise with Positive Trend:

- Both `peak_levels` and `MEM` compress all the levels into rather high values. This occurs because high-rate 
observations remain consistently elevated across all three seasons, causing these methods to overlook the remaining 
observations. 

2. Data with Noise and Positive Trend:

- All three methods exhibit higher burden levels, indicating that they successfully capture the exponentially 
increasing trend between seasons.

3. Data with Noise and Negative Trend:

- As observations exponentially decrease between seasons (with the highest observation this season being 3,735), 
we expect the burden levels to be lower.  This expectation is met across all three methods. However, the weighting 
of seasons in `intensity_levels` and `peak_levels` leads to older seasons having less impact on the burden levels as 
we progress forward in time. On the other hand, `MEM` includes all high-rate observations from the previous 10 seasons
without diminishing the importance of older seasons, which results in sustained very high burden levels.

- Notably, in the `MEM` method, the epidemic thresholds are positioned above the medium burden level. 
This means that the epidemic period begins only when the burden reaches the range of highest-rate observations 
observed in previous seasons.


This concludes, that using the `peak_levels` and `MEM` methods does not allow us to assess the burden before the season 
reaches the range of high-rate observations from previous seasons. In contrast, the intensity_levels method allows for 
continuous monitoring of the burden of current observation rate throughout the entire season.
